## Long Memory Analysis: Fractional Integration

### Motivation

The ACF of log(VIX) exhibits extremely slow decay, remaining significant for over 400 lags. This pattern is inconsistent with both:
- **Stationary AR processes** (exponential decay)
- **Unit root processes** (no decay)

Instead, this suggests **long-range dependence** or **long memory**, characteristic of fractionally integrated processes. While our ARIMA(1,1,1) model provides a practical approximation, we explore fractional differencing to explicitly model this phenomenon.

### Theory: ARFIMA Models

An ARFIMA(p, d, q) model allows for **fractional differencing** where $d \in \mathbb{R}$:

$$(1-B)^d x_t = \text{ARMA}(p,q)$$

The fractional difference operator is defined by the binomial expansion:

$$(1-B)^d = \sum_{k=0}^{\infty} \pi_k(d) B^k$$

where $\pi_0(d) = 1$ and $\pi_k(d) = \frac{(k-1-d)}{k}\pi_{k-1}(d)$ for $k \geq 1$.

**Key properties**:
- $d \in (0, 0.5)$: Stationary with long memory
- $d \in [0.5, 1)$: Non-stationary but mean-reverting
- $d = 1$: Unit root (standard differencing)
- $d > 1$: Non-stationary, non-invertible

### Estimation of d

```{r fracdiff-estimate}
library(fracdiff)

# Demean the series
log_vix_dm <- log_vix - mean(log_vix, na.rm = TRUE)

# Estimate d using Haslett-Raftery MLE
vix_fd <- fracdiff(log_vix_dm, nar = 0, nma = 0, M = 30)

# Results
fd_results <- data.frame(
  Parameter = c("d (fractional difference)", "Standard Error"),
  Value = c(vix_fd$d, vix_fd$stderror.dpq)
)

kable(fd_results,
      digits = 4,
      caption = "Fractional Differencing Parameter Estimate",
      booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position")
```

The estimated $\hat{d} \approx$ `r round(vix_fd$d, 3)` falls in the range $(0, 0.5)$, confirming that VIX exhibits **stationary long memory**. This means:

- Shocks decay **hyperbolically** (slowly) rather than exponentially
- The series is stationary but highly persistent
- Neither $d=0$ nor $d=1$ perfectly captures the dynamics

### ARFIMA(1, d, 0) Estimation

```{r arfima-ar1}
# Estimate ARFIMA with AR(1) component
vix_fd_ar1 <- fracdiff(log_vix_dm, nar = 1, nma = 0, M = 30)

# Display results
arfima_params <- data.frame(
  Parameter = c("d", "AR(1)", "SE(d)", "SE(AR1)"),
  Estimate = c(
    vix_fd_ar1$d,
    vix_fd_ar1$ar,
    vix_fd_ar1$stderror.dpq[1],
    vix_fd_ar1$stderror.dpq[2]
  )
)

kable(arfima_params,
      digits = 4,
      caption = "ARFIMA(1, d, 0) Parameter Estimates",
      booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position")
```

### Fractional Differencing Weights

The weights $\pi_k(d)$ determine how past observations are weighted in the fractionally differenced series:

```{r fracdiff-weights, fig.cap="Fractional Differencing Weights", fig.height=3.5}
# Compute weights
d_use <- vix_fd$d
p <- rep(1, 31)
for (k in 1:30) {
  p[k + 1] <- (k - d_use) * p[k] / (k + 1)
}

# Plot
plot(1:30, p[-1], 
     ylab = expression(pi(d)), 
     xlab = "Lag k", 
     type = "h",
     main = paste0("Fractional Differencing Weights (d = ", round(d_use, 3), ")"),
     col = "darkblue",
     lwd = 2)
abline(h = 0, col = "red", lty = 2)
```

The weights decay slowly, reflecting the long memory in the process.

### Whittle Likelihood Estimation

As an alternative to maximum likelihood, we employ the Whittle approximation based on the periodogram:

```{r whittle-estimation}
# Prepare series
series <- log_vix_dm
d0 <- 0.1
n.per <- nextn(length(series))
m <- floor(n.per / 2) - 1

# Compute periodogram
per <- Mod(fft(series - mean(series))[-1])^2 / n.per
g <- 4 * (sin(pi * ((1:m) / n.per))^2)

# Whittle log-likelihood
whit.like <- function(d) {
  g.d <- g^d
  sig2 <- sum(g.d * per[1:m]) / m
  log.like <- m * log(sig2) - d * sum(log(g)) + m
  return(log.like)
}

# Optimize
est <- optim(d0, whit.like, method = "L-BFGS-B", 
             hessian = TRUE, lower = -0.5, upper = 0.5)

# Extract results
d_whittle <- est$par
se_whittle <- 1 / sqrt(est$hessian)

whittle_results <- data.frame(
  Method = c("Haslett-Raftery MLE", "Whittle Likelihood"),
  `d estimate` = c(vix_fd$d, d_whittle),
  `Std Error` = c(vix_fd$stderror.dpq, se_whittle),
  check.names = FALSE
)

kable(whittle_results,
      digits = 4,
      caption = "Comparison of d Estimation Methods",
      booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position")
```

Both methods yield consistent estimates of $d$, confirming the robustness of the long memory finding.

### Residual Comparison: ARIMA vs ARFIMA

```{r residual-comparison, fig.cap="ACF Comparison: ARIMA vs ARFIMA Residuals", fig.height=5}
# Compute residuals
res_arima <- residuals(arima(log_vix, order = c(1, 1, 1)))
res_fd <- diffseries(log_vix, vix_fd$d)

# ACF plots
par(mfrow = c(2, 1))

acf(res_arima, 
    100, 
    xlim = c(4, 97), 
    ylim = c(-0.15, 0.15), 
    main = "ACF: ARIMA(1,1,1) Residuals",
    na.action = na.pass)

acf(res_fd, 
    100, 
    xlim = c(4, 97), 
    ylim = c(-0.15, 0.15), 
    main = "ACF: ARFIMA(0,d,0) Residuals",
    na.action = na.pass)

par(mfrow = c(1, 1))
```

**Observations**:

- Both models successfully remove most autocorrelation
- ARFIMA captures long memory explicitly through fractional d
- ARIMA approximates long memory through AR(1) and first-differencing
- Remaining autocorrelation in both suggests conditional heteroskedasticity (motivating GARCH)

### Summary Statistics Comparison

```{r model-comparison-stats}
# Compute statistics
resid_comparison <- data.frame(
  Model = c("ARIMA(1,1,1)", "ARFIMA(0,d,0)", "ARFIMA(1,d,0)"),
  d = c("1 (fixed)", 
        round(vix_fd$d, 4), 
        round(vix_fd_ar1$d, 4)),
  `Resid SD` = c(
    sd(res_arima, na.rm = TRUE),
    sd(res_fd, na.rm = TRUE),
    sd(diffseries(log_vix, vix_fd_ar1$d), na.rm = TRUE)
  ),
  `Resid Skewness` = c(
    moments::skewness(res_arima, na.rm = TRUE),
    moments::skewness(res_fd, na.rm = TRUE),
    moments::skewness(diffseries(log_vix, vix_fd_ar1$d), na.rm = TRUE)
  ),
  `Resid Kurtosis` = c(
    moments::kurtosis(res_arima, na.rm = TRUE),
    moments::kurtosis(res_fd, na.rm = TRUE),
    moments::kurtosis(diffseries(log_vix, vix_fd_ar1$d), na.rm = TRUE)
  ),
  check.names = FALSE
)

kable(resid_comparison,
      digits = 4,
      caption = "Residual Properties: ARIMA vs ARFIMA Models",
      booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position")
```

### Interpretation and Implications

The estimated $\hat{d} \approx$ `r round(vix_fd$d, 3)` provides important insights:

1. **Long Memory Confirmed**: VIX exhibits genuine long-range dependence, not just high AR persistence
2. **Stationarity**: Since $d < 0.5$, the series is covariance-stationary despite slow ACF decay
3. **Shock Persistence**: Volatility shocks decay at rate $O(k^{-(1+d)})$ rather than geometric decay
4. **Model Choice**: 
   - ARFIMA is theoretically correct for capturing long memory
   - ARIMA(1,1,1) provides a parsimonious practical approximation
   - Both yield similar residual behavior

**Practical Implications**:

- Forecasts from fractionally integrated models have **slower convergence** to the mean than ARIMA
- Long memory implies that **distant past observations** still influence current values
- This persistence is crucial for **risk management** - volatility shocks have lasting effects
- ARFIMA may provide **better long-horizon forecasts** than ARIMA

**Why We Proceed with ARIMA(1,1,1)**:

Despite the theoretical appeal of ARFIMA:
1. ARIMA(1,1,1) is more widely used and easier to interpret
2. Both models leave similar residual patterns (conditional heteroskedasticity)
3. GARCH modeling (next section) will address remaining volatility clustering
4. Computational simplicity favors ARIMA for this analysis

The key insight—that VIX has long memory—has been established and will inform our interpretation of results.
